\sectionthree{Multiplication: Karatsuba algorithm}
\begin{python0}
from solutions import *; clear()
\end{python0}
Multiplying two $n$-digit integers via the classical grade-school method takes $O(n^2)$ single-digit multiplications. Karatsuba’s insight reduces this to approximately $O(n^{1.585})$.


\begin{itemize}
  \item Let $x$ and $y$ be two nonnegative integers each of at most $n$ digits in base $T$ (e.g., $T=10$ or $T=2^{16}$).
  \item Write
    \[x = a\,T^{m} + b, \quad y = c\,T^{m} + d,\]
    where $m = \lceil n/2 \rceil$, and $a,b,c,d$ are roughly half-size.
  \item Observe the product:
    \[x\,y = a\,c\,T^{2m} + (a\,d + b\,c)\,T^{m} + b\,d.\]
  \item Instead of computing the four products $ac$, $ad$, $bc$, and $bd$, Karatsuba shows we need only three:
    \begin{enumerate}
      \item $A = a\,c$
      \item $B = b\,d$
      \item $C = (a + b)\,(c + d)$
    \end{enumerate}
  \item Then $a\,d + b\,c = C - A - B$.  Recombine:
    \[x\,y = A\,T^{2m} + (C - A - B)\,T^{m} + B.
    \]
  \item Each recursion halves the digit-length, leading to the recurrence
    \[T(n) = 3\,T\bigl(\lceil n/2\rceil\bigr) + O(n).\]
\end{itemize}

2. Complexity Analysis
\begin{itemize}
  \item Solve $T(n) = 3T(n/2) + cn$ by the Master Theorem:
    $$T(n) = \Theta\bigl(n^{\log_{2}3}\bigr) \approx \Theta(n^{1.585}).$$
  \item For large $n$, this outperforms $\Theta(n^2)$.  In practice, a cutoff to classical multiplication is used for small sizes to minimize overhead.
  \item Typical base threshold: when $n \le 32$ (or machine-word size), revert to $O(n^2)$ multiply.
\end{itemize}

3. Pseudocode
\begin{verbatim}
function Karatsuba(x, y):
    if x < T or y < T:
        return x * y            # base-case single-digit
    m = ceil(max(len(x), len(y)) / 2)
    (a, b) = split_at(x, m)
    (c, d) = split_at(y, m)
    A = Karatsuba(a, c)
    B = Karatsuba(b, d)
    C = Karatsuba(a + b, c + d)
    return A * T^(2*m) + (C - A - B) * T^m + B
\end{verbatim}

4. Threshold 
\begin{itemize}
  \item \emph{Threshold Choice:} Recursive overhead outweighs savings below a certain digit count.  Benchmark on target hardware.
  \item \emph{Base Case Multiply:} For $n < n_0$, perform schoolbook $O(n^2)$ multiply directly.
  \item \emph{Balanced Splits:} Use $m = \lceil n/2 \rceil$ to handle odd lengths.
  \item \emph{Efficient Addition:} Recombining requires only two additions and one subtraction of $O(n)$ cost.
  \item \emph{Radix Selection:} Use a large base $T=2^{16}$ or $2^{32}$ if implementing on binary machines to reduce recursion depth.
\end{itemize}

5. Code Examples

Python
\begin{verbatim}
def karatsuba(x, y):
    # Base case: single-digit
    if x < 10 or y < 10:
        return x * y
    # Determine split size m
    n = max(len(str(x)), len(str(y)))
    m = (n + 1) // 2
    # Split x and y
    high1, low1 = divmod(x, 10**m)
    high2, low2 = divmod(y, 10**m)
    # Three recursive calls
    A = karatsuba(high1, high2)
    B = karatsuba(low1, low2)
    C = karatsuba(high1 + low1, high2 + low2)
    # Combine results
    return A * 10**(2*m) + (C - A - B) * 10**m + B
\end{verbatim}



\input{exercises/rsa-30/main.tex}
\input{exercises/rsa-31/main.tex}
\input{exercises/rsa-32/main.tex}
\input{exercises/rsa-33/main.tex}
