\chapter{Basic number theory}

\boxpar{
\textsc{Suggestions}.
For this chapter, state the basic axioms and properties/theorems of $\Z$.
Provide proofs. 
But remember that most of the properties/theorems can be generlized
to properties/theorems for rings.
It's still a good idea to prove the facts for $\Z$ since $\Z$ is not
as abstract as general rings and will prepare you for the general results.
}

The area of Number theory is huge. We will only cover number theory until we reach prime factorization. The reason being that one of the most important ciphers that is used in the world is based on the difficulty of factorizing two very large prime numbers. SPOLER ALERT IT's RSA.

Since, we now know that the study of number theory is huge, it is also important to know that many different fields of mathematics have advanced because of nubmer theory such as automorphic theory, theory of modular forms, algebraic geometry etc.

The term ``elementary'' here does not mean that the content is easy rather that is actually the begining of the number thoery. Since, it will take a very long time to begin from nothing and cover everything in number thoery, we will start directly from the study of $\Z$ and it's properties.


We need to think of $Z$ as not just a set in itself, but rather the set including operations $+$, $*$, $0$, $1$

\newpage\input{group.tex}
\newpage\input{ring.tex}
\newpage\input{axioms-of-Z.tex}
\newpage\input{divisibility.tex}
\newpage\input{congruences.tex}
\newpage\input{euclidean-property.tex}
\newpage\input{extended-euclidean-algorithm.tex}
\newpage\input{gcd.tex}
\newpage\input{prime.tex}
\newpage\input{euler-totient-function.tex}
%\newpage\input{relations.tex}

%\newpage\input{congruence-classes.tex}

