\sectionthree{Linear feedback shift register}
\begin{python0}
from solutions import *; clear()
\end{python0}

LFSRs are efficient mechanisms for generating pseudorandom bit sequences. They operate by using the current state to deterministically calculate the next bit in a sequence.

\subsection*{Basic Concept}
\begin{enumerate}
  \item Start with an initial bit sequence (seed).
  \item Apply a linear function (XOR) to selected bits.
  \item Shift the register and append the new bit.
\end{enumerate}

\subsection*{Mathematical Foundation}
Exclusive-OR (XOR) in modulo 2 arithmetic:
\[
0 \oplus 0 = 0,\quad 0 \oplus 1 = 1,\quad 1 \oplus 0 = 1,\quad 1 \oplus 1 = 0.
\]

\subsection*{Example}
Consider a 4-bit LFSR with initial state $\mathbf{x}^{(0)}=(0,1,1,1)$ and feedback rule:
\[
x_{n+4} = x_n \oplus x_{n+2} \pmod{2}.
\]
Compute:
\[
\begin{aligned}
x_4 &= x_0 \oplus x_2 = 0 \oplus 1 = 1,\\
x_5 &= x_1 \oplus x_3 = 1 \oplus 1 = 0,\\
x_6 &= x_2 \oplus x_4 = 1 \oplus 1 = 0,\\
x_7 &= x_3 \oplus x_5 = 1 \oplus 0 = 1,\\
x_8 &= x_4 \oplus x_6 = 1 \oplus 0 = 1,\\
\end{aligned}\]
sequence continues similarly. The output stream begins:
\[
0111\,1001\,1001\,1\dots
\]
This register has period 7 (less than the maximum $2^4-1=15$).

\subsection*{Applications in Cryptography}
\begin{itemize}
  \item Generate pseudorandom sequences efficiently with minimal resources.
  \item Implementable in hardware or software.
  \item Certain feedback polynomials yield maximal period $2^n-1$.
  \item Vulnerable if the structure and output bits are known (solvable via linear algebra).
\end{itemize}
\input{exercises/lfsr-0/main.tex}
\input{exercises/lfsr-1/main.tex}
\input{exercises/lfsr-2/main.tex}
\input{exercises/lfsr-3/main.tex}
\input{exercises/lfsr-4/main.tex}
\input{exercises/lfsr-5/main.tex}
