\sectionthree{One-time pad cipher}
\begin{python0}
from solutions import *; clear()
\end{python0}
\section*{One-Time Pad}

The one-time pad (OTP) offers perfect secrecy (information-theoretic security) when used correctly. Its key properties are:

\begin{itemize}
  \item Use a truly random key stream, as long as the plaintext.
  \item Encrypt by bitwise XOR: 
    \[
      C_i = P_i \oplus K_i.
    \]
  \item Every possible plaintext of that length is equally likely given the ciphertext.
  \item An attacker with only the ciphertext gains zero information about the plaintext.
  \item Security does not rely on computational assumptions—no amount of computing power can break it.
\end{itemize}

\subsection*{Key Requirements}
\begin{itemize}
  \item The key must be generated from a truly random source.
  \item The key length must be at least as long as the message.
  \item Each key bit is used exactly once and never reused.
  \item The key must remain perfectly secret.
\end{itemize}

\subsection*{Practical Considerations}
\begin{itemize}
  \item Generating and distributing large, truly random keys is difficult.
  \item Secure key exchange and storage are the main challenges.
  \item In practice, OTPs are rare—often replaced by stream ciphers that approximate randomness.
\end{itemize}

\input{exercises/otp-0/main.tex}

There's a rumor that during the Cold War, Washington D.C. communicates with
Moscow using one time pad.

(Using the concept of entropy of information theory, Claude Shannon can prove that
the ciphertext contains no information about the plaintext, other than the length.)



